---
title: "Searching for potential Alzheimer’s disease biomarkers through vervet biomarker and behavior analysis, Cindy project1"
author: "Tingting"
date: "12/22/2021"
output: html_document
---

```{r}
library(xlsx)
library(tidyverse)
library(tidyr)
library(lme4)
```

## Data cleaning
Load biomarker data
```{r}
data <- read.xlsx('../input/LEMERE VERVET CSF-STANFORD RAW DATA_TZ.xls', header = T, sheetIndex = 1, check.names=FALSE)
```

```{r}
rownames(data) <- data$Exp
data <- data[,-1]  
```

Data transpose
```{r}
data_t <- as.data.frame(t(data))
data_t$vervet_time <- rownames(data_t)
data_t <- data_t[, c(616, 1:615)]
```

Load cohort data
```{r}
df_cohort <- read.xlsx('../input/Communicome cohort GB RP CL 12-17-2021_TZ.xlsx', header = T, sheetIndex = 7, check.names=FALSE)
colnames(df_cohort) <- c("ID", "vervet_time", "Sex", "Age", "Cognition")
```

Add age group
```{r}
df_cohort$Age_group <- factor(cut(df_cohort$Age, breaks=c(0, 10, 15, 20, 100), labels=c("young", "middle", "middle old", "very old")))

df_cohort$ID <- as.factor(df_cohort$ID)
df_cohort$vervet_time <- as.factor(df_cohort$vervet_time)
df_cohort$Sex <- as.factor(df_cohort$Sex)
df_cohort$Age_group <- as.factor(df_cohort$Age_group)
df_cohort$Cognition <- as.factor(df_cohort$Cognition)
```

Merge two sets of data
```{r}
df_merged <- merge(df_cohort, data_t, by=c("vervet_time", "vervet_time")) 
```

Clean merged data
```{r}
names(df_merged) <- gsub(" ", "_", names(df_merged))
names(df_merged) <- gsub(x = names(df_merged), pattern = "\\(", replacement = "")  
names(df_merged) <- gsub(x = names(df_merged), pattern = "\\)", replacement = "")  
names(df_merged) <- gsub(x = names(df_merged), pattern = "\\-", replacement = "_")  
```


# Q1

Explore: Linear regression of biomarker ~ ID + Age + Sex
```{r}
df <- df_merged

ICAM2.age.sex.int.regression <- lm(ICAM2 ~ ID + Age + Sex, df)
summary(ICAM2.age.sex.int.regression)
```

Extract age slop prob of linear regressions between biomarker ~ ID + Age + Sex
```{r}
df <- df_merged

lm_multi_matrix <- matrix(0,length(df)-6,5)
rownames(lm_multi_matrix) <- colnames(df[7: length(df)])
reg_ls=c()
for(i in colnames(df[7: length(df)])){
  lm_reg <-  paste(i, "~ ID", "+ Age", "+ Sex")
  reg_ls <- lm(lm_reg, df)
  lm_multi_matrix[i,1] <- paste(i, "~ ID", "+ Age", "+ Sex")
  lm_multi_matrix[i,2] <- tryCatch(unname(1 - pf(summary(reg_ls)$fstatistic[1], summary(reg_ls)$fstatistic[2], summary(reg_ls)$fstatistic[3])), error = function(e) NA)
  lm_multi_matrix[i,3] <- tryCatch(summary(reg_ls)$adj.r.squared, error = function(e) NA)
  lm_multi_matrix[i,4] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-1, 1], error = function(e) NA)
  lm_multi_matrix[i,5] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-1, 4], error = function(e) NA)
  }

# rename lm_multi_matrix
colnames(lm_multi_matrix) <- c("Linear regression", "F p-value", "Adjusted R-squared", "Age slope", "Age slope Pr")
lm_multi_matrix <- as.data.frame(lm_multi_matrix)
lm_multi_matrix[ , colnames(lm_multi_matrix[2: length(lm_multi_matrix)])] <- apply(lm_multi_matrix[ , colnames(lm_multi_matrix[2: length(lm_multi_matrix)])], 2, function(x) as.numeric(x))

# add adjusted p-values
lm_multi_matrix$`Fdr Age slope Pr` <- p.adjust(lm_multi_matrix$`Age slope Pr`, method = "fdr")

lm_multi_matrix$`Bonferroni Age slope Pr` <- p.adjust(lm_multi_matrix$`Age slope Pr`, method = "bonferroni")

write.csv(lm_multi_matrix, file="../output/1_Lm_Biomarker_AgeMultiPredictor.csv")

# save Age pro significant
lm_multi_matrix_sig <- subset(lm_multi_matrix, lm_multi_matrix$`Age slope Pr` < 0.05)
write.csv(lm_multi_matrix_sig, file="../output/1_Lm_Biomarker_AgeMultiPredictor_sig.csv")
```

Scatterplot: biomarker vs. age for age pro significant biomarkers, taking intercept and age slope from biomarker ~ ID + Age + Sex
```{r}
# slope is affected by ID
df_sig <- lm_multi_matrix_sig

pdf(file = "../output/1_Scatterplot_Biomarker_Age_Sig.pdf", width = 8, height = 6) 

for(i in rownames(df_sig)){
  p = ggplot(df, aes(Age, df[[i]], color=Sex)) + xlab("Age") + ylab(i) +
    theme(axis.text=element_text(size=12), axis.title=element_text(size=14, face="bold")) +
    geom_point() + 
    geom_abline(intercept = coef(lm(df[[i]] ~ ID + Age + Sex , df))[1], slope = coef(lm(df[[i]] ~ ID + Age + Sex , df))[33], col='red', size=1.5) +
    annotate("text", x=Inf, y=Inf, hjust=1, vjust=1, label =paste0("Age slope: ", coef(lm(df[[i]] ~ ID + Age + Sex , df))[33]) )
  print(p)
  }

dev.off()
```

# Q2: Linear regression of biomarker ~ Age + ID + Sex + Age:Cognition, (Cognition=ORT_global)

Explore: Linear regression of biomarker ~ ID + Age + Sex + Age:Cognition
```{r}
df <- df_merged

ICAM2.age.sex.int.regression <- lm(ICAM2 ~ ID + Age + Sex + Age:Cognition, df)
summary(ICAM2.age.sex.int.regression)
```

Extract age slop prob of linear regressions between biomarker ~ ID + Age + Sex + Age:Cognition
```{r}
df <- df_merged

lm_multi_matrix <- matrix(0,length(df)-6,11)
rownames(lm_multi_matrix) <- colnames(df[7: length(df)])
reg_ls=c()
for(i in colnames(df[7: length(df)])){
  lm_reg <-  paste(i, "~ ID", "+ Age", "+ Sex", "+ Age:Cognition")
  reg_ls <- lm(lm_reg, df)
  lm_multi_matrix[i,1] <- paste(i, "~ ID", "+ Age", "+ Sex", "+ Age:Cognition")
  lm_multi_matrix[i,2] <- tryCatch(unname(1 - pf(summary(reg_ls)$fstatistic[1], summary(reg_ls)$fstatistic[2], summary(reg_ls)$fstatistic[3])), error = function(e) NA)
  lm_multi_matrix[i,3] <- tryCatch(summary(reg_ls)$adj.r.squared, error = function(e) NA)
  lm_multi_matrix[i,4] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-3, 1], error = function(e) NA)
  lm_multi_matrix[i,5] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-3, 4], error = function(e) NA)
  lm_multi_matrix[i,6] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-2, 1], error = function(e) NA)
  lm_multi_matrix[i,7] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-2, 4], error = function(e) NA)
  lm_multi_matrix[i,8] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-1, 1], error = function(e) NA)
  lm_multi_matrix[i,9] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients)-1, 4], error = function(e) NA)
  lm_multi_matrix[i,10] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients), 1], error = function(e) NA)
  lm_multi_matrix[i,11] <- tryCatch(summary(reg_ls)$coefficients[nrow(summary(reg_ls)$coefficients), 4], error = function(e) NA)
  }

# rename lm_multi_matrix
colnames(lm_multi_matrix) <- c("Linear regression", "F p-value", "Adjusted R-squared", "Age slope", "Age slope Pr", "Sexm slope", "Sexm slope Pr", "Age:CognitionSolidly bad slope", "Age:CognitionSolidly bad slope Pr", "Age:CognitionSolidly good slope", "Age:CognitionSolidly good slope Pr")
lm_multi_matrix <- as.data.frame(lm_multi_matrix)
lm_multi_matrix[ , colnames(lm_multi_matrix[2: length(lm_multi_matrix)])] <- apply(lm_multi_matrix[ , colnames(lm_multi_matrix[2: length(lm_multi_matrix)])], 2, function(x) as.numeric(x))

# add adjusted p-values
lm_multi_matrix$`Fdr Age slope Pr` <- p.adjust(lm_multi_matrix$`Age slope Pr`, method = "fdr")
lm_multi_matrix$`Fdr Sexm slope Pr` <- p.adjust(lm_multi_matrix$`Sexm slope Pr`, method = "fdr")
lm_multi_matrix$`Fdr Age:CognitionSolidly bad slope Pr` <- p.adjust(lm_multi_matrix$`Age:CognitionSolidly bad slope Pr`, method = "fdr")
lm_multi_matrix$`Fdr Age:CognitionSolidly good slope Pr` <- p.adjust(lm_multi_matrix$`Age:CognitionSolidly good slope Pr`, method = "fdr")


lm_multi_matrix$`Bonferroni Age slope Pr` <- p.adjust(lm_multi_matrix$`Age slope Pr`, method = "bonferroni")
lm_multi_matrix$`Bonferroni Sexm slope Pr` <- p.adjust(lm_multi_matrix$`Sexm slope Pr`, method = "bonferroni")
lm_multi_matrix$`Bonferroni Age:CognitionSolidly bad slope Pr` <- p.adjust(lm_multi_matrix$`Age:CognitionSolidly bad slope Pr`, method = "bonferroni")
lm_multi_matrix$`Bonferroni Age:CognitionSolidly good slope Pr` <- p.adjust(lm_multi_matrix$`Age:CognitionSolidly good slope Pr`, method = "bonferroni")

write.csv(lm_multi_matrix, file="../output/2_Lm_Biomarker_AgeMultiPredictor.csv")

# save Age pro significant
lm_multi_matrix_sig <- subset(lm_multi_matrix, lm_multi_matrix$`Age slope Pr` < 0.05)
write.csv(lm_multi_matrix_sig, file="../output/2_Lm_Biomarker_AgeMultiPredictor_sig.csv")
```

Scatterplot for one: biomarker vs. age for age pro significant biomarkers, taking intercept and age slope in ORT global group from biomarker ~ ID + Age + Sex + Age:Cognition
```{r}
df <- df_merged

coefSum <- coef(lm(ICAM2 ~ ID + Age + Sex + Age:Cognition, df))

ggplot(df, aes(Age, ICAM2, color=Cognition, shape=Sex)) + xlab("Age") + ylab(i) +
  theme(axis.text=element_text(size=12), axis.title=element_text(size=14, face="bold")) +
  geom_point() + 
  geom_abline(intercept = coefSum[1], slope = coefSum[33], col='#F8766D', size=1.5) +
  geom_abline(intercept = coefSum[1], slope = coefSum[35] + coefSum[33], col='#7CAE00', size=1.5) +
  geom_abline(intercept = coefSum[1], slope = coefSum[36] + coefSum[33], col='#619CFF', size=1.5) +
  annotate("text", x=Inf, y=Inf, hjust=1, vjust=1, color="#F8766D", label =paste0("Slope of cognition declining: ", coefSum[33])) +
  annotate("text", x=Inf, y=Inf, hjust=1, vjust=3, color="#7CAE00", label =paste0("Slope of cognition solidly bad: ", coefSum[35] + coefSum[33])) +
  annotate("text", x=Inf, y=Inf, hjust=1, vjust=5, color="#619CFF", label =paste0("Slope of cognition solidly good: ", coefSum[36] + coefSum[33]) )

```

Scatterplot for all: biomarker vs. age for age pro significant biomarkers, taking intercept and age slope in ORT global group from biomarker ~ ID + Age + Sex + Age:Cognition
```{r}
df <- df_merged
#df_sig <- lm_multi_matrix_sig

pdf(file = "../output/2_Scatterplot_Biomarker_Age_ORTgroup.pdf", width = 8, height = 6) 

for(i in colnames(df[7: length(df)])){
  coefSum <- coef(lm(df[[i]] ~ ID + Age + Sex + Age:Cognition, df))
  p = ggplot(df, aes(Age, df[[i]], color=Cognition, shape=Sex)) + xlab("Age") + ylab(i) +
    theme(axis.text=element_text(size=12), axis.title=element_text(size=14, face="bold")) +
    geom_point() + 
    geom_abline(intercept = coefSum[1], slope = coefSum[33], col='#F8766D', size=1.5) +
    geom_abline(intercept = coefSum[1], slope = coefSum[35] + coefSum[33], col='#7CAE00', size=1.5) +
    geom_abline(intercept = coefSum[1], slope = coefSum[36] + coefSum[33], col='#619CFF', size=1.5) +
    annotate("text", x=Inf, y=Inf, hjust=1, vjust=1, color="#F8766D", label =paste0("Slope of cognition declining: ", coefSum[33])) +
    annotate("text", x=Inf, y=Inf, hjust=1, vjust=3, color="#7CAE00", label =paste0("Slope of cognition solidly bad: ", coefSum[35] + coefSum[33])) +
    annotate("text", x=Inf, y=Inf, hjust=1, vjust=5, color="#619CFF", label =paste0("Slope of cognition solidly good: ", coefSum[36] + coefSum[33]) )
  print(p)
  }

dev.off()
```


# Q3 ANOVA, Biomarker on Cognition groups

Input file
```{r}
df <- df_merged
```

Boxplot of an example biomarker on Cognition group
```{r}
#pdf(file = "../output/BoxplotBiomarkerORTgGroup.pdf", width = 6, height = 6)
boxplot(df$A2M_hu ~ droplevels(df$Cognition), xlab = "ORT global groups", ylab = "A2M_hu", cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
#dev.off()
```

Boxplot of individual biomarker on ORT global groups for all
```{r include=FALSE}
pdf(file = "../output/3_Boxplot_Biomarker_ORTgGroup.pdf", width = 6, height = 6)

for(i in colnames(df[7: length(df)])){
  boxplot(df[[i]] ~ droplevels(df$Cognition), xlab = "ORT global groups", ylab = i, cex.lab=1.5, cex.axis=1, cex.main=1.5, cex.sub=1.5, outline=FALSE)
  stripchart(df[[i]] ~ droplevels(df$Cognition), method = "jitter", pch = 19, col = 2:4, vertical = TRUE, add = TRUE)
  }

dev.off()
```

ANOVA between ORT global groups for an individual biomarker
```{r}
anovatest <- aov(df$A2M_hu~ df$Cognition)
summary(anovatest)
```


Loop through biomarker data for ANOVA test on ORT global groups
```{r}
anova_matrix <- matrix(0,length(df)-6,1)
rownames(anova_matrix) <- colnames(df[7: length(df)])
for(i in colnames(df[7: length(df)])){
 if(length(unique(df[["Cognition"]][!is.na(df[[i]])])) < 3){ # ort score groups should have at least 3 levels. If not, assign NA to that i=biomarker
   anova_matrix[i, ] <- NA
   next
 }
probability=summary(aov(df[[i]] ~ df[["Cognition"]]))[[1]][["Pr(>F)"]][1]
if(!is.null(probability))
  anova_matrix[i,1] <- summary(aov(df[[i]] ~ df[["Cognition"]]))[[1]][["Pr(>F)"]][1]
}

# rename and output
colnames(anova_matrix) <- c("ANOVA ORT global group Pr")
anova_matrix <- as.data.frame(anova_matrix)

# add adjusted p-value
anova_matrix$Fdr <- p.adjust(anova_matrix$`ANOVA ORT global group Pr`, method = "fdr" )
anova_matrix$Bonferroni <- p.adjust(anova_matrix$`ANOVA ORT global group Pr`, method = "bonferroni" )
write.csv(anova_matrix, file="../output/3_Biomarker_ORTgGroup_Anova.csv")
```


# Q3 t-test, Biomarker on Cognition groups

Explore t-test between two selected ORT global groups for an individual biomarker
```{r}
ttest <- t.test(df$A2M_hu[df$Cognition=="Declining"], df$A2M_hu[df$Cognition=="Solidly bad"])
names(ttest)
ttest$p.value
```


Loop through biomarker data for two sample t-test on ORT global groups
```{r}
ttest_matrix <- matrix(0,length(df)-6, 3)
rownames(ttest_matrix) <- colnames(df[7: length(df)])
for(i in colnames(df[7: length(df)])){
    if(length(unique(df[df$Cognition=="Declining" & !is.na(df[[i]]), c(i)])) >= 2 & length(unique(df[df$Cognition=="Solidly bad" & !is.na(df[[i]]), c(i)])) >= 2) #ttest in each group must have at least two unique numbers
      ttest_matrix[i,1] <-  t.test(as.numeric(df[df$Cognition=="Declining", c(i)]), as.numeric(df[df$Cognition=="Solidly bad", c(i)]))$p.value
    else
      ttest_matrix[i,1] <- NA
    
    if(length(unique(df[df$Cognition=="Declining" & !is.na(df[[i]]), c(i)])) >= 2 & length(unique(df[df$Cognition=="Solidly good" & !is.na(df[[i]]), c(i)])) >= 2) #ttest in each group must have at least two unique numbers
      ttest_matrix[i,2] <-  t.test(as.numeric(df[df$Cognition=="Declining", c(i)]), as.numeric(df[df$Cognition=="Solidly good", c(i)]))$p.value
    else
      ttest_matrix[i,2] <- NA

    if(length(unique(df[df$Cognition=="Solidly bad" & !is.na(df[[i]]), c(i)])) >= 2 & length(unique(df[df$Cognition=="Solidly good" & !is.na(df[[i]]), c(i)])) >= 2) #ttest in each group must have at least two unique numbers
      ttest_matrix[i,3] <-  t.test(as.numeric(df[df$Cognition=="Solidly bad", c(i)]), as.numeric(df[df$Cognition=="Solidly good", c(i)]))$p.value
    else
      ttest_matrix[i,3] <- NA
}

# rename table
colnames(ttest_matrix) <- c("P-value of Declining vs. Solidly bad", "P-value of Declining vs. Solidly good", "P-value of Solidly bad vs. Solidly good")
ttest_matrix <- as.data.frame(ttest_matrix)

# add adjusted p-values
ttest_matrix$`Fdr Declining vs. Solidly bad` <- p.adjust(ttest_matrix$`P-value of Declining vs. Solidly bad`, method = "fdr")
ttest_matrix$`Fdr Declining vs. Solidly good` <- p.adjust(ttest_matrix$`P-value of Declining vs. Solidly good`, method = "fdr")
ttest_matrix$`Fdr Solidly bad vs. Solidly good` <- p.adjust(ttest_matrix$`P-value of Solidly bad vs. Solidly good`, method = "fdr")


ttest_matrix$`Bonferroni Declining vs. Solidly bad` <- p.adjust(ttest_matrix$`P-value of Declining vs. Solidly bad`, method = "bonferroni")
ttest_matrix$`Bonferroni Declining vs. Solidly good` <- p.adjust(ttest_matrix$`P-value of Declining vs. Solidly good`, method = "bonferroni")
ttest_matrix$`Bonferroni Solidly bad vs. Solidly good` <- p.adjust(ttest_matrix$`P-value of Solidly bad vs. Solidly good`, method = "bonferroni")

# save output
write.csv(ttest_matrix, file="../output/3_Biomarker_ORTgGroup_tTest.csv")
```

